Pomace: A Grappa for Non-Volatile Memory \vspace{-.2in}

### 2012-12-13 doc/nvm2013/paper.tex
edits and more squeezing

\FloatBarrier
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{../asplos13/figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
%\vspace{-6ex}
\end{center}
\end{figure}

@@

\item [Tasking system.] Our tasking system supports lightweight
multithreading to tolerate communication latency and global distributed
workstealing (i.e., tasks can be stolen from any node in the system), which
provides automated load balancing.  Grappa thus tolerates many
microseconds of latency by multiplexing thousands of tasks per x86 core.

\item[Communication layer.] Modern commodity networks
support high bandwidth only for large messages and irregular applications
frequently issue small requests.   Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million remote references per second per node.  Aggregation is largely invisible to the application programmer.

\item[Distributed shared memory.] Our \textls*{\small DSM} system provides support for
fine-grain access to data anywhere in the system. It supports synchronization
operations on global data, explicit local caching of any memory in the system,
and support for operations on remote data (delegating operations to owner node).
By tight integration with the tasking system and the
communication layer, our \textls*{\small DSM} system offers high aggregate random-access bandwidth to remote data.

@@

Emerging non-volatile memory technologies offer an alternative strategy:
place the entire data set on \textls*{\small SSD} or \textls*{\small PCM};
use Grappa to tolerate the latency directly, much in the same
way as it is used now to tolerate network latency, or indirectly when transferring blocks of working-set
data into \textls*{\small RAM}; save energy per available byte;
recover from faults faster with most of the data preserved in non-volatile memory.

@@

\section{Applying non-volatile memory to Grappa}
The natural point of integration for non-volatile memories in Grappa
is to store the global heap in nonvolatile memory, since the heap
usually contains data that is infrequently modified. All accesses to
this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer. If the
access is to local memory, Grappa issues a prefetch for the address
and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
the access is to memory stored on another node, Grappa issues a
message to that node requesting the data and context switches to other
work until the reply arrives. Non-volatile memory would be accessed in
the same way: If a request is to memory stored in non-volatile memory,
Grappa would initiate the request and context switch to other work
until the request had completed.

\subsection{Challenges in using commodity \textls*{\small SSD}\hspace{0.75pt}s}
Grappa's goal is to provide good irregular application performance on
commodity hardware. Unfortunately, commodity non-volatile memories are
difficult to use in this way for three reasons:

@@

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
perform on the order of 1 million random reads per second, and many
are limited to less than 100K/s. In contrast, cores may be idle in
Grappa if each node cannot make 100 million random reads per second of
the global heap. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}.

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
as normal disks: accesses must occur through the kernel, incurring a
high context-switch overhead. Grappa depends on user-mode access to
the network layer. This capability has been explored in recent research~\cite{caulfield:2012}.

@@

For simplicity, we omit discussion of write and erase operations. We
assume a non-volatile Grappa would store frequently written temporary
data in \textls*{\small DRAM} and would use its standard latency-tolerance techniques
to cover the cost of infrequent writes.

\subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
A natural first step would be to add one or more \textls*{\small SSD}\hspace{0.75pt}s to each Grappa
cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
accesses and context switch to tolerate the latency of \textls*{\small SSD}
access. Unfortunately this is unlikely to work well, for two
reasons. First, the \textls*{\small SSD} must accept requests at a rate sufficient to
keep the cores busy. Grappa's target is for each cluster node to
generate 100 million random 8-byte requests per second, but most \textls*{\small SSD}\hspace{0.75pt}s
can accept requests at a rate two to four orders of magnitude slower.
Second, the \textls*{\small SSD} must be able to support many requests operating in
parallel. Little's Law suggests that if requests take 20 $\mu$s for
the \textls*{\small SSD} to service, and if the node generates requests at a rate of
100 million per second, the \textls*{\small SSD}\hspace{0.75pt}s must support 2000 concurrent
requests. Again, existing \textls*{\small SSD}\hspace{0.75pt}s support orders of magnitude less request concurrency.

@@

%\paragraph{Scaling up the Flash array}
\noindent{\bf Scaling up the Flash array}
Existing Flash-based \textls*{\small SSD}\hspace{0.75pt}s are designed for large block requests. An
\textls*{\small SSD} for Grappa could be designed to support high concurrency, small
requests. Where a traditional \textls*{\small SSD} might put multiple Flash chips on
the same bus, or use Flash chips in parallel to build a wider,
high-bandwidth bus, a Grappa \textls*{\small SSD} would try to maximize parallelism by
putting each chip on its own bus. Serial Flash might be used to
minimize pin count requirements, at the cost of some additional
latency. This design could scale to request rates, but the challenge
would be chip count: to support 100M reads per second, an array of
20$\mu$s flash would require 2000 chips. This is possible, but expensive.

\paragraph{Scaling down the processors.}
%\noindent{\bf Scaling down the processors.}
Alternatively, rather than scaling up the flash array, processors
could be scaled down to match the flash request rate. A low-power
processor with few cores might generate only 1 million references per
second. This could be satisfied with only 20\,x\,20$\mu$s Flash chips.

There are two problems with this design for our purpose. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency~\cite{feo:xmt}.

@@

While similar techniques have provided benefit for traversal-oriented applications in
restricted situations~\cite{magt:2010}, they depend on a sufficient request
density to each flash page to get a benefit. As datasets scale up, this
density may decrease, limiting the benefit of this technique.

For iterative applications where the entire graph is regularily
visited, an windowed approach similar to~\cite{graphchi:osdi12} may
provide more benefit. In this approach, graph data is constantly
streamed between SSD and DRAM. Requests wait to execute until the
relevant fragment of the graph is in memory. Request density is not a
concern for these applications; instead, bandwidth between the SSD and
DRAM is the limiting factor.

@@

\paragraph{\textls*{\small PCM} as Main Memory.}
%\noindent{\bf \textls*{\small PCM} as Main Memory.}
Much recent research has explored the use of \textls*{\small PCM} as a main memory
replacement. Grappa would easily be able to take advantage of such a
development. When a Grappa program references low-locality data,
a prefetch request is issued for that data a few cycles before the
thread is context switched in. A \textls*{\small PCM}-based
Grappa would do the same, but would issue the prefetches earlier to
cover the longer latency of \textls*{\small PCM}.

\paragraph{\textls*{\small PCM} as attached storage.}
%\noindent{\bf \textls*{\small PCM} as attached storage.}
Grappa could also support a \textls*{\small PCM} array in a separate address space (for
instance, on the \textls*{\small PCIe} bus). \textls*{\small PCM} chips available today~\cite{micronPCM} support a
64-bit read in 200 ns; this means that 20 chips acting independently
would be sufficient to reach our target request rate.

Just as in network communications, issuing individual \textls*{\small PCIe} bus
operations for each memory request has too high overhead to get a
reasonable request rate.  Instead, Grappa would delay threads
performing operations in the \textls*{\small PCM} array until it could send a buffer
with enough requests to amortize the overhead. The \textls*{\small PCM} array would
perform the operations and buffer the results. When a sufficient
amount have completed, the \textls*{\small PCM} array would write the results to \textls*{\small DRAM}\@,
and one core would deliver the results to the requesting threads,
waking them.

>>>>

\FloatBarrier
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{../asplos13/figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \nocaptionrule\caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
\vspace{-5ex}
\end{center}
\end{figure}

@@

\item [Tasking system.] Our tasking system supports lightweight
multithreading to tolerate communication latency; it also supports global distributed
workstealing (i.e., tasks can be stolen from any node in the system), which
provides automated load balancing.  Grappa thus tolerates many
microseconds of latency by multiplexing thousands of tasks per x86 core.

\item[Communication layer.] Modern commodity networks
support high bandwidth only for large messages, but irregular applications
frequently issue small requests.   Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million 8-byte remote references per second per node.  Aggregation is largely invisible to the application programmer.

\item[Distributed shared memory.] Our \textls*{\small DSM} system provides support for
fine-grain access to data anywhere in the system. It supports synchronization
operations on global data, explicit local caching of any memory in the system,
and support for operations on remote data (delegating operations to the data's owner node).
By tight integration with the tasking system and the
communication layer, our \textls*{\small DSM} system offers high aggregate random-access bandwidth to remote data.

@@

Emerging non-volatile memory technologies offer an alternative strategy:
place the entire data set on Flash or \textls*{\small PCM}, and
%% use Grappa to tolerate the latency either directly, much in the same
%% way as it is used now to tolerate network latency, or indirectly when transferring blocks of working-set
%% data into \textls*{\small RAM}. This could save energy per available byte and allow
%% faster fault recovery with most of the data preserved in non-volatile memory.
use Grappa to tolerate the access latency. This could save energy per available byte and allow
faster fault recovery with most of the data preserved in non-volatile memory.

@@

\section{Applying non-volatile memory to Grappa}
The natural point of integration for non-volatile memories in Grappa
is to store the global distributed shared heap in nonvolatile memory, since for irregular applications the heap
often holds data that is infrequently modified. All accesses to
this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer. 
If an 
access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address
and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
the access is to memory stored on another node, Grappa issues a
message to that node requesting the data and context switches to other
work until the reply arrives. Non-volatile memory would be accessed in
the same way: If a request is to memory stored in non-volatile memory,
Grappa would initiate the request and context switch to other work
until the request had completed.

\subsection{Challenges in using commodity \textls*{\small SSD}\hspace{0.75pt}s}
Grappa's goal is to provide good irregular application performance on
mass market hardware. Unfortunately, commodity \textls*{\small SSD}\hspace{0.75pt}s are
difficult to use in this way for three reasons:

@@

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
perform on the order of 1 million random reads per second, and many
are limited to less than 100K/s. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, cores may be idle in
Grappa if each node cannot make 100 million random reads per second of
the global heap.

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
as normal disks: accesses must occur through the kernel, incurring a
high context-switch overhead. Grappa depends on user-mode access to
the network layer. This approach has been applied to \textls*{\small SSD}\hspace{0.75pt}s in recent research~\cite{caulfield:2012}.

@@

For simplicity, we omit discussion of write and erase operations. We
assume a non-volatile Grappa would store frequently written temporary
data in \textls*{\small DRAM} and would use its standard latency-tolerance techniques
to cover the cost of infrequent writes to non-volatile memory.

\subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
A natural first step would be to add one or more mass market \textls*{\small SSD}\hspace{0.75pt}s to each Grappa
cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
accesses and context switch to tolerate the latency of \textls*{\small SSD}
access. Unfortunately this is unlikely to work well, for two
reasons. First, as discussed earlier, the \textls*{\small SSD} must accept requests at a rate sufficient to
keep the cores busy; most \textls*{\small SSD}\hspace{0.75pt}s
can accept requests at a rate two to four orders of magnitude slower than required.
Second, the \textls*{\small SSD} must be able to support many requests operating in
parallel. Little's Law suggests that if requests take 25 $\mu$s for
the \textls*{\small SSD} to service, and if the node generates requests at a rate of
100 million per second, the \textls*{\small SSD}\hspace{0.75pt}s must support 2500 concurrent
requests. Again, existing \textls*{\small SSD}\hspace{0.75pt}s support orders of magnitude less request concurrency.

@@

\paragraph{Scaling up the Flash array.}
%\noindent{\bf Scaling up the Flash array}
Existing Flash-based \textls*{\small SSD}\hspace{0.75pt}s are designed for large block requests. An
\textls*{\small SSD} for Grappa could be designed to support high concurrency, small
requests. Where a traditional \textls*{\small SSD} might put multiple Flash chips on
the same bus, or use Flash chips in parallel to build a wider,
high-bandwidth bus, a Grappa \textls*{\small SSD} would try to maximize parallelism by
putting each chip on its own bus. Serial Flash might be used to
minimize pin count requirements, at the cost of some additional
latency. This design could scale to request rates, but the challenge
would be chip count: to support 100 million reads per second, an array of
25$\mu$s flash would require 2500 chips. This is possible, but expensive.

\paragraph{Scaling down the processors.}
%\noindent{\bf Scaling down the processors.}
Alternatively, rather than scaling up the flash array, processors
could be scaled down to match the flash request rate. A low-power
processor with few cores might generate only 1 million references per
second. This could be satisfied with only 25\,x\,25$\mu$s Flash chips.
There are two problems with this design for our purpose. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.

@@

While similar techniques have provided benefit for graph traversal applications in
restricted situations~\cite{magt:2010}, they depend on a sufficient request
density to each flash page to get a benefit. As datasets scale up and machines scale out, this
density may decrease, limiting the benefit of this technique.

For iterative applications where the entire graph is regularily
visited, an windowed approach similar to~\cite{graphchi:osdi12} may
provide more benefit. In this approach, graph data is constantly
streamed between \textls*{\small SSD} and \textls*{\small DRAM}. Requests wait to execute until the
relevant fragment of the graph is in memory. Request density is not a
concern for these applications; instead, bandwidth between the \textls*{\small SSD} and \textls*{\small DRAM}
is the limiting factor.

@@

\paragraph{\textls*{\small PCM} as Main Memory.}
%\noindent{\bf \textls*{\small PCM} as Main Memory.}
Recent research has explored the use of \textls*{\small PCM} as a main memory
replacement~\cite{pcm:isca09}. Grappa would easily be able to take advantage of such a
development. When a Grappa program references low-locality data,
a prefetch request is issued for that data a few cycles before the
thread is context switched in. A \textls*{\small PCM}-based
Grappa would do the same, but would issue the prefetches earlier to
cover the longer latency of \textls*{\small PCM}.

\paragraph{\textls*{\small PCM} as attached storage.}
%\noindent{\bf \textls*{\small PCM} as attached storage.}
Grappa could also transparently support a \textls*{\small PCM} array in a separate address space (for
instance, on the \textls*{\small PCIe} bus). \textls*{\small PCM} chips available today~\cite{micronPCM} support a
64-bit read in 200 ns; this means that 20 chips acting independently
would be sufficient to reach our target request rate.

Just as in network communications, issuing individual \textls*{\small PCIe} bus
operations for each memory request has too high overhead to get a
reasonable request rate.  Instead, Grappa would delay threads
performing operations in the \textls*{\small PCM} array until it could send a buffer
with enough requests to amortize the overhead. The \textls*{\small PCM} array would
perform the operations and buffer the results. When a sufficient
number have completed, the \textls*{\small PCM} array would write the results to \textls*{\small DRAM}\@,
and one core would deliver the results to the requesting threads,
waking them.
